Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
be a bit more explicit about properties of page sharing. + a note.
Formatting only

TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in section \ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing
JIT\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

Overall, we believe that disabling it on all benchmarks except the
performance benchmarks in section \ref{sec:performance-bench} is better
because we can minimise non-determinism. We also do not want to depend
on the capabilities of the JIT in these experiments.

@@

In figure \ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

For the results in figure \ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$13\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$13\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
more iterations per second than a single thread.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in figure \ref{fig:performance-nojit}.

@@

The results are presented in figure \ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. In general, we see that the group of
non-locked benchmarks certainly scales best. The other three scale
barely or not at all with the number of threads. The slowdown factor
from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
performance in half of the benchmarks.

>>>>

TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
figure~\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing
JIT\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

Overall, we believe that disabling it on all benchmarks except the
performance benchmarks in section~\ref{sec:performance-bench} is better
because we can minimise non-determinism. We also do not want to depend
on the capabilities of the JIT in these experiments.

@@

In figure~\ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

For the results in figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$13\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$13\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
more iterations per second than a single thread.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in figure~\ref{fig:performance-nojit}.

@@

The results are presented in figure~\ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. In general, we see that the group of
non-locked benchmarks certainly scales best. The other three scale
barely or not at all with the number of threads. The slowdown factor
from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
performance in half of the benchmarks.
